---
title: 'Pratical work: EM algorithm'
author: "Thomas Roiseux"
date: "`r Sys.Date()`"
output: pdf_document
---
\part{Simulation}
We first generate:
\begin{enumerate}
\item A sample of \(n=100\) observations with a Poisson law using \(\lambda=3\).
\item A sample of \(n=200\) observations with a Poisson law using \(\lambda=15\).
\item A vector of \(300\) coordinates, which the 100 first are 1 and the others are 2.
\end{enumerate}
```{r generatingPoisson}
sample3 <- rpois(100, 3)
print(sample3)
sample15 <- rpois(200, 15)
print(sample15)
v <- c()
for (i in 1:100)
{
  v <- c(v, 1, recursive = TRUE)
}
for (i in 1:200)
{
  v <- c(v, 2, recursive = TRUE)
}
```

Now, we are going to generate a Poisson law using two components:
```{r poisson2Comp}
#Settings constants
pi1 <- 0.4
pi2 <- 0.6
lambda1 <- 3
lambda2 <- 15
sample1 <- rpois(300, lambda1)
sample2 <- rpois(300,lambda2)

mixed_sample = pi1 * sample1 + pi2 * sample2
print(mixed_sample)
```
\part{EM-algorithm}
We first want the initialization of the EM-algorithm: