---
title: 'Pratical work: EM algorithm'
author: "Thomas Roiseux"
date: "`r Sys.Date()`"
output: pdf_document
---
\part{Simulation}
We first generate:
\begin{enumerate}
\item A sample of \(n=100\) observations with a Poisson law using \(\lambda=3\).
\item A sample of \(n=200\) observations with a Poisson law using \(\lambda=15\).
\item A vector of \(300\) coordinates, which the 100 first are 1 and the others are 2.
\end{enumerate}
```{r generatingPoisson}
sample3 <- rpois(100, 3)
print(sample3)
sample15 <- rpois(200, 15)
print(sample15)
v <- c()
for (i in 1:100)
{
  v <- c(v, 1, recursive = TRUE)
}
for (i in 1:200)
{
  v <- c(v, 2, recursive = TRUE)
}
```

Now, we are going to generate a Poisson law using two components:
```{r poisson2Comp}
#Settings constants
pi1 <- 0.4
pi2 <- 0.6
lambda1 <- 3
lambda2 <- 15
sample1 <- rpois(300, lambda1)
sample2 <- rpois(300,lambda2)

mixed_sample = pi1 * sample1 + pi2 * sample2
print(mixed_sample)
```
\clearpage
\part{EM-algorithm}
\section{Initialization}
We first want the initialization of the EM-algorithm:
```{r em_init}
pi1 <- 0.4
pi2 <- 0.6
lambda1 <- 3
lambda2 <- 15

dpois2comp <- function(x){
  return(pi1 * exp(-lambda1) /
           factorial(x) * lambda1 ** x +
           pi2 * exp(-lambda2) / 
           factorial(x) * lambda2 ** x)
}
```
We will also use the \verb|dpois2comp| function to have the density
of the Poisson law with two components.
\section{Estimation}
We now want to compute all the expectancies, before finding the element that will maximize it:
```{r eme}
em_e = function(data, theta){
  result <- c()
  
  return(result)
}
```
\section{Maximisation}
Having all the possible values for the expectancy, we now want the element that maximizes the expectancy:
```{r emm}
em_m = function(expectancy){
  i <- 1
  for (j in 2:length(expectancy)){
    if (expectancy[i] < expectancy[j]) {
      i <- j
    }
  }
  return(i)
}
```
\section{Global algorithm and application}
Using all the previous functions, we are going to write the EM-algorithm:
```{r em}
em <- function(data, k = NULL){
  i <- 0
  theta <- runif(n = 1, min = 0, max = 50)
  theta2 <- -1
  while (theta != theta2 || (!(is.null(k)) && i < k)){
    theta2 <- theta
    i <- i + 1
    expec <- em_e(data, theta)
    theta <- em_m(expec)
  }
  return(theta)
}
```
We can now use our algorithm on the previous data set:
```{r ex}
print(em(mixed_sample))
```